---
title: "Assignment 5: Data Visualization"
author: "Kendall Fitzgerald"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
# loading packages required for this assignment
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
library(ggplot2)

# assigning a variable to processed data location
processed_data = "./Data/Processed_KEY"

# reading in data
PeterPaul.chem.nutrients <- read.csv(
  here(
    processed_data,"NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"),
  stringsAsFactors = TRUE)

NIWO.litter <- read.csv(
  here(processed_data,"NEON_NIWO_Litter_mass_trap_Processed.csv"),
  stringsAsFactors = TRUE)

#2 
# fix format of dates
PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
NIWO.litter$collectDate <- ymd(NIWO.litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

custom_theme <- theme_gray() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top")
theme_set(custom_theme)
  

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
Phos.plot <- PeterPaul.chem.nutrients %>%
  ggplot(aes(x = tp_ug, y = po4,color = lakename))+
  geom_point()+
  ylim(0,50)+
  geom_smooth(method = "lm", se = FALSE)


Phos.plot +
  labs(title = "Total Phosphorus by Phosphate Levels in Peter the & Paul Lakes",
       y = "Phosphate",
       x = "Phosphorus")+
  theme(
    plot.title = element_text(size = 7),
    legend.title = element_blank()
  )

  

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
#creating boxplot with temperature as y-axis
boxplot_temp <- 
  ggplot(PeterPaul.chem.nutrients, aes(x= month,
                                       y=temperature_C,
                                       color=lakename))+
  geom_boxplot()

boxplot_temp1 <-
  boxplot_temp + 
  labs(y="Temperature",
       x="Month")+
  theme(legend.title = element_blank())
  

#creating boxplot with TP as y-axis
boxplot_tp <- 
  ggplot(PeterPaul.chem.nutrients, aes(x= month,
                                       y=tp_ug,
                                       color=lakename))+
  geom_boxplot()

boxplot_tp1 <-
  boxplot_tp + 
  labs(y="TP",
       x="Month")


#creating boxplot with TN as y-axis
boxplot_tn <- 
  ggplot(PeterPaul.chem.nutrients, aes(x= month,
                                       y=tn_ug,
                                       color=lakename))+
  geom_boxplot()

boxplot_tn1 <-
  boxplot_tn + 
  labs(y="TN",
       x="Month")
    

#combining boxplots together and fixing aesthetics
cowplot1 <- plot_grid(
  boxplot_tn1 + theme(legend.position = "none"),
  boxplot_tp1+ theme(legend.position = "none"),
  boxplot_temp1+ theme(legend.position = "none"), 
  align = 'vh',
  hjust = -1,
  nrow = 1)

#extracting legend from one of boxplots
legend <- get_legend(
  boxplot_temp1+ 
     guides(color = guide_legend(nrow = 1))+
     theme(legend.position = "right")
)

#creating a new cowplot with new legend position
cowplot2 <-
  plot_grid(cowplot1, legend, ncol = 1, rel_heights = c(1, .1))

#creating a new cowplot with an aesthetic title 
cowplot3 <-
  cowplot2 +
  labs(title = "Comparison of Temp, TN, and TP Across Peter & Paul Lakes")+
  theme(
     plot.title = element_text(size = 10.25, face = "bold")
  )
  

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6


#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: